---
title: 'DS202: Lab 5'
author: "MyTien Kien; kmytien"
date: "4/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(ggplot2)
library(tidyr)
library(classdata)
library(lubridate)
library(dplyr)

accident <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)

person <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```

## **Questions**

**1. Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?**
```{r}
q1 <- accident %>%
      group_by(DAY_WEEK) %>%
      summarise(Accidents = n()) %>%
      mutate(DAY_WEEK = as.factor(DAY_WEEK))

ggplot(q1, aes(x = DAY_WEEK, y = Accidents)) + geom_bar(stat = 'identity') + 
      scale_x_discrete(labels = c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday')) +
      ggtitle("Day v.s. Total Accidents") + xlab("Day of the Week") + ylab("Number of Accidents")
```

Say something here


**2. Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)**
```{r}
#only fatal injury or fatal injury and more?
q2 <- person %>% filter(INJ_SEV == 4)

head(q2)
```

Say something here


**3. Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)**
```{r}
#need to handle missing variables
q3 <- person %>%
      filter(INJ_SEV == 4) %>%
      group_by(STATE, MAKE) %>%
      summarise(Accidents = n()) %>%
      arrange(desc(Accidents))

drop_na(q3)
```

Say something here


**4.Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)**
```{r}

```


**5. Join the accident and person table (work out which variable(s) to use)**
```{r}
#does it depend on the certain join?
#couldn't join by state for some reason so did st_case
acc_person <- accident %>% inner_join(person, by = "ST_CASE")
```


**6. Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results and explain what you find.**
```{r}

```


**7. Now plot a choropleth map of the number of deaths on a county level. Also explain what you find.**
```{r}

```


**8. Is summer or winter more dangerous? Does this depend on states? Explore and explain.**
```{r}

```

